---
title: "p8105_hw2_ef2721"
author: "Erfan Faridmoayer"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

```{r}
library(tidyverse)
library(readxl)
```


## Problem 1

Below we import and clean data from `NYC_Transit_Subway_Entrance_And_Exit_Data.csv`. First data is imported, then the `Route` column names that are not `dbl` are turned into `chr`. We then use janitor function, select the noted columns, and change `entry` from `yes` / `no` to a logical variable. 

```{r}
trans_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>%
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

The presentation is not yet "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables. 

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
trans_df %>% 
  select(station_name, line) %>% 
  distinct
```

There are 465 unique stations based on the code chunk above. 

To identify unique stations that are `ada` complaints, we first filter the data from ada that are `TRUE`. Then, similar to above, we select based on station name and line. Below is the code chunk explanation.

```{r}
trans_df %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

There are 84 compliant stations.

For the next sub-question, we first filter the stations with no vending machine. We then see how many stations from that subset allow entry to be possible. Mean is proportion here since entry is a logical variable. 

```{r}
trans_df %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

the proportion desired was 0.38.

We use the `pivot_longer` function to reformat the data to add route number and route name as distinct variables. Subsequently, we use the `filter` function to narrow down which stations service A line, and of those which are ADA compliant, respectively. The code chunk below addresses that 

```{r}
trans_df %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct
```

There are 60 distinct stations that serve the A line

```{r}
trans_df %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

There are 17 distinct stations which serve the A line and are ADA compliant.

## Problem 2

The `readxl` library has been introduced earlier in the code. In this problem, we first read the data, narrowing to the sheet desired. To avoid unnecessary images, the data were selected to the numerical table. We then use the `janitor` function to clean the data, and selected dumpster-specific data. Using the `filter` function, the rows with summary values were removed. Lastly, the `as.integer` function was used to round sports_balls values to the nearest round number. Below, is the code chunk implementing above:

```{r}
trash_df=
  read_excel("./data/Trash-Wheel-Collection-Totals-7-2020-2.xlsx", 
      sheet = "Mr. Trash Wheel", 
      range = "A2:N535") %>% 
  janitor::clean_names() %>% 
  select(-homes_powered) %>% 
  drop_na %>% 
  mutate(
    sports_balls = as.integer(sports_balls)
  )
```

